\documentclass[letterpaper, 11pt]{article}
\usepackage{/Users/miles/Documents/latex/syllabus}

%% Metadata
\renewcommand{\Title}{Physics 9C Lab Syllabus}
\renewcommand{\Instructor}{Miles Moser}
\renewcommand{\Email}{mmoser@ucdavis.edu}
\renewcommand{\Hours}{Mondays 3--4 pm, RESSLR 158}

\begin{document}
\setlength{\headheight}{14pt}
\insertTitle

\textbf{Description.} The lab portion of 9C is where you, with the help of your classmates, construct your own understanding of physical concepts by actually testing them out. Generally, you will spend the first part of the lab actually doing the experiments and the second writing up an informal report to turn in at the end. You will be in small groups of two or three, and turn in one single report as a group. I will shuffle the groups randomly two times this quarter, so I can better judge your individual contributions.

\textbf{Preparation.} You should prepare for each lab by reading the manual in advance. All of the manuals, along with some other useful information, are combined into one file located here:

\begin{center}
    physics.ucdavis.edu/Classes/Physics9Lab/Phy9CLab/Phy9CLabWinter18.pdf
\end{center}

\textbf{Grades.} Your grade will depend on your attendance, lab reports, and participation. Attendance will determine the maximum grade you can receive:
\begin{table}[H]
    \centering
    \begin{tabular}{ll}
        \textbf{Attendance Record} & \textbf{Highest Attainable Grade} \\
        \toprule
        No missed labs or one excused miss & High pass \\
        One unexcused miss or two excused misses & Pass \\
        Two misses, at least one of them unexcused & Low pass \\
        Three or more misses of any kind & No pass \\
        \bottomrule
    \end{tabular}
\end{table}

Your lab report grades and participation will then determine whether you get the highest grade for your attendance level or something lower. Most of you will get a pass. A high pass requires exceptional effort.

\textbf{Makeup Labs.} We only have eight labs this quarter, so it’s very important to make up any lab that you miss. The best way to do it is in another one of my sections, but a different TA’s section is fine, too. If you go to another TA’s lab, make sure that they know that you are making up my lab so they can confirm with me that you did. Check in with me afterwards to make sure I gave you credit for making it up. Makeups are kind of an administrative hassle, so only two are allowed for the quarter. Here is a list of my sections this quarter:
\begin{table}[H]
    \centering
    \begin{tabular}{llll}
        \textbf{Section} & \textbf{Day} & \textbf{Time} & \textbf{Room} \\
        \toprule
        C03 & Tuesday & 1:10 -- 3:30 pm & RESSLR 162 \\
        C07 & Wednesday & 8:00 -- 10:20 am & RESSLR 162 \\
        A05 & Thursday & 8:00 -- 10:20 am & RESSLR 162 \\
        \bottomrule
    \end{tabular}
\end{table}

\textbf{Expectations.} Have an open mind, and be ready to abandon your prior knowledge in favor of rebuilding an idea from scratch. Don’t be condescending. Treat each other with respect, but do not hesitate to share your ideas. Most importantly, try to have a good time doing physics. In this class, you are in control, so make the best of it.

\end{document}