\documentclass{article}

%% Formatting
\usepackage[letterpaper,margin=1.5in]{geometry} % page setup
\usepackage[shortlabels]{enumitem}  % customizeable enumerators
\usepackage[USenglish]{babel}   % ensure correct hyphenation
\usepackage[T1]{fontenc}        % validate output font
\usepackage[utf8]{inputenc}     % validate input characters
\usepackage{siunitx}    % SI units
\usepackage{graphicx}   % include graphics
\usepackage{booktabs}   % tables
\usepackage{float}      % [H] option for floats
\usepackage{parskip}    % remove paragraph indentations

%% Content
\usepackage{amsmath}    % math
\usepackage{physics}    % physics

%% Metadata
\newcommand{\Title}     {Problem Set 1}
\newcommand{\DueDate}   {January 25, 2018}
\newcommand{\Course}    {PHY 215B}

\begin{document}
{\huge\bf\Title}

Due \DueDate \hfill \Course

%------------------------------------------------------------------------------
\hrulefill

\textbf{Problem 1.} If operators $A$ and $B$ do not commute, show that
\begin{equation*}
    e^{A+B} = e^Ae^Be^{(-1/2)[A,B]},
\end{equation*}
where $[A,B]$ is a commutator. Discuss this result in your own terms.
 
%------------------------------------------------------------------------------
\hrulefill

\textbf{Problem 2.} Show that if $A$ and $B$ are Hermitian operators, then all of the following operators are Hermitian:
\begin{enumerate}[(a)]
    \item $A+B$
    \item $AB+BA$
    \item $i(AB-BA)$
    \item $A^nB^m + B^mA^n$, where $m$ and $n$ are integers
    \item $(AB)^\dagger = B^\dagger A^\dagger$
\end{enumerate}

%------------------------------------------------------------------------------
\hrulefill

\textbf{Problem 3.} Show that $AA^\dagger$ is Hermitian even if $A$ is not.

%------------------------------------------------------------------------------
\hrulefill

\textbf{Problem 4.} Given a truncated harmonic oscillator centered at the origin:
\begin{equation*}
    H = 
    \begin{cases} 
        \dfrac{p^2}{2m} + \dfrac{kx^2}{2} & |x| \leq a, \\[1em]
        \dfrac{p^2}{2m} & a \leq |x| \leq \ell, \,\, a < \ell,
    \end{cases}
\end{equation*}
and a wave function of the form:
\begin{equation*}
    \Psi = b_1\exp\left(\frac{i\pi x}{\ell}\right) + b_2\exp\left(\frac{2i\pi x}{\ell}\right) + b_3\exp\left(\frac{3i\pi x}{\ell}\right),
\end{equation*}

set up a $3\times 3$ matrix of $H$ using the functions given in $\Psi$. $b_i$ are not normalization constants. If the integrations are too hard, you should just set up the matrix correctly. If $\Psi$ is expanded onto three lowest energy basis functions of the un-truncated harmonic oscillator, set up the matrix for the transformation between the two sets of functions. Is the matrix unitary? If it is not, form a unitary matrix.

%------------------------------------------------------------------------------
\hrulefill

\textbf{Problem 5.} For the one-dimensional harmonic oscillator problem, show that:
\begin{enumerate}[(i)]
     \item its energy is positive definite.
     \item the difference between neighboring states is $\hbar\sqrt{k/m}$ using Heisenberg's approach of quantum mechanics.
 \end{enumerate} 

\end{document}