\documentclass[11pt]{article}
\usepackage{/Users/miles/Documents/latex/hw}

%% Metadata
\renewcommand{\Title}{RIOT Data Reflection}
\renewcommand{\Course}{PHY 371}
\renewcommand{\Date}{November 16, 2017}
\renewcommand{\Author}{Miles Moser}

\begin{document}
\insertTitle

My RIOT data wasn't surprising, and I thought it mostly reflected a normal day. I thought the only part that wasn't representative was the time spent "Not interacting," which was higher than usual because I didn't prepare that DL as much as I normally do. My time spent with each group was fairly uniform, with less time spent on the groups that usually do well.

I do think that the data is fairly illuminating about my teaching preferences. In general, I try to be hands-off for the first half of a section, letting the students explore various avenues, right or wrong. Then, I try to wrap things up with a short lecture, correcting misconceptions and keeping Q\&A to a minimum in the interest of time. This is reflected in the data, with about 40\% of my time spent observing and about 33\% of my time spent explaining content, with small amounts of open dialogue in between.

When I took RIOT data for Matt, I noticed the fairly uniform way he pathed from group to group, going in a predictable order. This usually meant that, if he felt pressed for time on a section, he would skip the same group or groups every time. I realized that I do this, too, so I've tried to randomize my walking patterns and talk to groups in a different order for each section. I also noticed that, when his groups got stuck, they tended to just sit there and do nothing until Matt talked to them. Each time he talked to them, though, they would be able to keep working after he left. So, to prevent downtime for students, I try to talk to each group for a shorter amount of time, getting them started but not neglecting the other groups.

I found this activity really useful, but I think it was mostly because of the time observing Matt's section. I think that observation might have been just as or nearly as useful whether I was actively taking data or not.

\end{document}